\documentclass[11pt,a4paper]{article}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{natbib}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}

% --- Theorem environments ---
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}

% --- Macros ---
% \cola{} forces upright shape so it renders correctly inside theorem/lemma/
% definition bodies (which are italic by default; T1/.../m/scit is undefined).
\newcommand{\cola}{\textup{\textsc{Cola}}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}

% Review-flag macro for items that warrant Prof. Highley's confirmation.
% Renders inline in red so flagged items are visually scannable.
% To remove all flags before submission: redefine to \newcommand{\highley}[1]{}
\newcommand{\highley}[1]{\textcolor{red}{\textbf{[Prof.~Highley:}~#1\textbf{]}}}

% --- Title ---
\title{Incentive Compatibility of Carry-Over Lottery Allocation:\\
A Manipulation Bound and Historical Comparison of the \cola{} Family}

\author{
  Timothy Highley\thanks{La Salle University. \texttt{highley@lasalle.edu}}
  \and
  Kaushik Rajan\thanks{Independent researcher. \texttt{kaushi@alumni.ncsu.edu}}
}

\date{\today}

\begin{document}

\maketitle

\begin{abstract}
The National Basketball Association (NBA) draft lottery creates a documented incentive misalignment: teams eliminated from playoff contention can improve their expected draft position by losing additional games. \citet{munro2020notanking} prove that no mechanism based on end-of-season records alone can simultaneously reward poor performance and eliminate this incentive. Carry-Over Lottery Allocation (\cola{}) sidesteps the impossibility by replacing single-season records with multi-year playoff history as the basis for draft priority \citep{highley2026cola}. We make two contributions to the \cola{} literature. First, we derive an analytic upper bound of approximately $4.0\%$ on the worst-case probability gain from pool manipulation under empirically calibrated index distributions. The bound confirms the simulation-based estimate of \citet{highley2026cola} from the analytic side and yields a closed-form sensitivity to the underlying parameters. Second, we present the first historical backtest of the \cola{} family across 26 NBA seasons (1999--2025), comparing five variants (Simple, Simple Lottery, Classic, Countdown, and Capped \cola{}) against 776 verified draft outcomes and 55 lottery-day-attributed pre-draft trades. The analysis identifies trade handling and lottery-day attribution as the two largest sources of implementation ambiguity, and quantifies the per-series tanking exposure introduced by Capped \cola{}'s stockpile bound. Together, the results support treating \cola{} as a family of mechanisms that admit multiple adoption trajectories rather than a single design.
\end{abstract}

% --- Sections ---
\input{sections/01-introduction}
\input{sections/02-background}
\input{sections/03-cola-family}
\input{sections/04-manipulation-bound}
\input{sections/05-backtesting}
\input{sections/06-comparison}
\input{sections/07-discussion}

% --- Bibliography ---
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
